On exponential sums arising from the classical doubling method

### 2025-02-07 paper.tex
Trying to reduce to Kondo's

We now use this to compute $\dblGaussSum{\tau}{\chi}$. We have that $$\dblGaussSum{\tau}{\chi} = q^{-k^2} \GaussSumScalar{\chi_{\GL_k}}{\fieldCharacter} \sum_{x, y \in \GL_k\left(\finiteField\right)} \fieldCharacter\left(\trace x\right) \chi^{-1}\left(\det y\right) \fieldCharacter\left(\trace y\right) \pi\left(-y^{-1} x\right).$$
The last equality implies $$\dblGaussSum{\tau}{\chi} = \centralCharacter{\tau}\left(-1\right) \GaussSum{\chi}{\fieldCharacter}^k \GaussSumScalar{\tau}{\fieldCharacter} \GaussSumScalar{\tau^{\vee} \times \chi^{-1}}{\fieldCharacter}.$$

@@

\subsubsection{Basic torus computation}
In this section we compute the doubling method Gauss sum for a torus of the form $\finiteFieldExtension{2m}^1$.

@@

Suppose that $T \cong \finiteFieldExtension{m}^1$, and let $t \in T$ be an element such that $t$ is contained only in the torus $T$. Then
$$K\left(t\right) = \sum_{\theta} \theta\left(t\right) \innerproduct{K}{\DeligneLusztigInduction{T}{G}\left(\theta\right)},$$
where the sum is over all the characters of $T$. Using the formula $$\dblGaussSumScalar{\pi}{\chi} = \tau\left( \theta' \times \chi^{-1}, \fieldCharacter_{2m} \right) \theta\left(-1\right) \tau\left(\chi^2, \fieldCharacter\right)^m,$$
we get
$$K\left(t\right) = \frac{\sizeof{\lieAlgebra}^{\frac{k}{2}} \cdot \DeligneLusztigInduction{T}{G}\left(\theta\right)\left(1\right)}{\sizeof{G}} \sum_{\theta} \theta\left( \left(-1\right)^k t\right) \prod_{j=1}^k \tau\left(\theta' \times \chi_j^{-1}, \fieldCharacter_{2m}\right) \tau\left(\chi_j^2, \fieldCharacter\right)^m.$$

>>>>

We now use this to compute $\dblGaussSum{\tau}{\chi}$. We have that $$\dblGaussSum{\tau}{\chi} = q^{-k^2} \GaussSumScalar{\chi_{\GL_k}}{\fieldCharacter} \sum_{x, y \in \GL_k\left(\finiteField\right)} \fieldCharacter\left(\trace x\right) \chi^{-1}\left(\det y\right) \fieldCharacter\left(\trace y\right) \pi\left(-y^{-1} x\right).$$
The last equality implies the following result.
\begin{theorem}\label{thm:gln-doubling-gauss-sum-in-terms-of-kondo}For any irreducible representation $\tau$ of $\GL_k\left(\finiteField\right)$ and any non-trivial character $\chi \colon \multiplicativegroup{\finiteField} \to \multiplicativegroup{\cComplex}$, the following identity holds:
	$$\dblGaussSumScalar{\tau}{\chi} = \centralCharacter{\tau}\left(-1\right) \GaussSumScalar{\chi}{\fieldCharacter}^k \GaussSumScalar{\tau}{\fieldCharacter} \GaussSumScalar{\tau^{\vee} \times \chi^{-1}}{\fieldCharacter}.$$
\end{theorem}

@@

\subsubsection{General linear groups computation}
Suppose that $F \colon \GL_k\left(\finiteField\right) \to \cComplex$ is a class function. If $\chi \colon \multiplicativegroup{\finiteField} \to \multiplicativegroup{\cComplex}$ is a non-trivial character, we define $$\dblVirtualGaussSumScalar{F}{\chi} = q^{-\frac{k^2}{2}} \sum_{g \in \GL_k\left(\finiteField\right)} \Phi_{\chi}\left(g\right) F\left(g\right).$$
It is clear that for an irreducible representation $\tau$ of $\GL_k\left(\finiteField\right)$, we have $$\dblVirtualGaussSumScalar{\trace \tau}{\chi} = \dblGaussSumScalar{\tau}{\chi} \cdot \dim \tau.$$
Since the assignment $g \mapsto \Phi_{\chi}\left(g\right)$ depends only on the semisimple part of $g$, by \cite[Theorem in Section 1.2]{SaitoShinoda2000} we have that if $T \cong \multiplicativegroup{\finiteFieldExtension{k}}$ and if $\theta \colon \multiplicativegroup{\finiteFieldExtension{k}} \to \multiplicativegroup{\cComplex}$ is a character then
$$\dblVirtualGaussSumScalar{R_{T, \theta}}{\chi} = \frac{\grpIndex{\GL_k\left(\finiteField\right)}{T}}{q^{\frac{k^2}{2}}} \sum_{\substack{t \in \multiplicativegroup{\finiteFieldExtension{k}}\\
t \ne 1}} \theta\left(t\right) \chi\left(\FieldNorm{k}{1}\left(t-1\right)\right).$$
Using \Cref{thm:gln-doubling-gauss-sum-in-terms-of-kondo} and the fact that Kondo's Gauss sum is the same for every subrepresentation of $R_{T, \theta}$, we have the equality $$\dblVirtualGaussSumScalar{R_{T, \theta}}{\chi} = \left(-1\right)^k \theta\left(-1\right) q^{-k} R_{T, \theta}\left(1\right) \tau\left(\chi^2, \fieldCharacter\right)^k \tau\left(\theta, \fieldCharacter\right) \tau\left(\theta^{-1} \times \chi^{-1}, \fieldCharacter\right).$$

\subsubsection{Split torus computation}
In this section we compute the doubling method Gauss sum for a torus of the form $\multiplicativegroup{\finiteFieldExtension{k}}$. Let $\theta \colon \multiplicativegroup{\finiteFieldExtension{k}} \to \multiplicativegroup{\cComplex}$ and $\chi \colon \multiplicativegroup{\finiteField} \to \multiplicativegroup{\cComplex}$ be characters such that $\chi^2 \ne 1$. Our goal is to compute $$\sum_{\substack{x \in \multiplicativegroup{\finiteFieldExtension{k}}\\
		x \ne 1}} \theta \left(x\right) \chi\left(\FieldNorm{k}{1}\left(x - 1\right)\right) \chi\left(\FieldNorm{k}{1}\left(x^{-1} - 1\right)\right).$$

This can be rewritten as
$$\chi\left(-1\right)^k \sum_{\substack{x \in \multiplicativegroup{\finiteFieldExtension{k}}\\
		x \ne 1}} \theta \left(x\right) \chi^{-1}\left(\FieldNorm{k}{1}\left(x\right)\right) \chi^2\left(\FieldNorm{k}{1}\left(x - 1\right)\right).$$
Denote by $\theta_{\chi^{-1}} \colon \multiplicativegroup{\finiteFieldExtension{k}} \to \multiplicativegroup{\cComplex}$ the character $$\theta_{\chi^{-1}}\left(x\right) = \theta\left(x\right) \chi^{-1}\left(\FieldNorm{k}{1}\left(x\right)\right).$$

Hence we get
$$\sum_{\substack{x \in \multiplicativegroup{\finiteFieldExtension{k}}\\
		x \ne 1}} \theta \left(x\right) \chi\left(\FieldNorm{k}{1}\left(x - 1\right)\right) \chi\left(\FieldNorm{k}{1}\left(x^{-1} - 1\right)\right) = \frac{q^{\frac{k^2}{2}}}{\grpIndex{\GL_k\left(\finiteField\right)}{T}} \dblVirtualGaussSumScalar{R_{T, \theta_{\chi^{-1}}}}{\chi^2},$$
which by the results of the previous section equals
%$$R_{T,\theta}\left(1\right) = \left(-1\right)^{k-1} \frac{\grpIndex{\GL_k\left(\finiteField\right)}{T}}{q^{\frac{k^2 - k}{2}}}$$
$$-\theta\left(-1\right) q^{-\frac{k}{2}} \tau\left(\chi^4, \fieldCharacter\right)^k \tau\left(\theta \times \chi^{-1}, \fieldCharacter\right) \tau\left(\theta^{-1} \times \chi^{-1}, \fieldCharacter\right).$$

\subsubsection{Elliptic torus computation}
In this section we compute the doubling method Gauss sum for a torus of the form $\finiteFieldExtension{2m}^1$.

@@

Suppose that $T \cong \finiteFieldExtension{2m}^1$, and let $t \in T$ be an element such that $t$ is contained only in the torus $T$. Then
$$K\left(t\right) = \sum_{\theta} \theta\left(t\right) \innerproduct{K}{\DeligneLusztigInduction{T}{G}\left(\theta\right)},$$
where the sum is over all the characters of $T$. Using the formula $$\dblGaussSumScalar{\pi}{\chi} = \tau\left( \theta' \times \chi^{-1}, \fieldCharacter_{2m} \right) \theta\left(-1\right) \tau\left(\chi^2, \fieldCharacter\right)^m,$$
we get
$$K\left(t\right) = \frac{\sizeof{\lieAlgebra}^{\frac{k}{2}} \cdot \DeligneLusztigInduction{T}{G}\left(\theta\right)\left(1\right)}{\sizeof{G}} \sum_{\theta} \theta\left( \left(-1\right)^k t\right) \prod_{j=1}^k \tau\left(\theta' \times \chi_j^{-1}, \fieldCharacter_{2m}\right) \tau\left(\chi_j^2, \fieldCharacter\right)^m.$$
